### 2017-04-27 paper/jfp-submission/evaluation.tex
prune orphans

\paragraph{Evaluation Goals}
%
There are three questions we seek to answer with our evaluation:
%
\begin{enumerate}
\item \emphbf{Witness Coverage}
      How many ill-typed programs can we find witnesses for?
\item \emphbf{Witness Complexity}
      How \emph{complex} are the traces produced by the witnesses?
\item \emphbf{Witness Utility}
      How \emph{helpful} (qualitatively and quantitatively)
      are the witnesses and traces in debugging type errors?
\end{enumerate}

@@

@@

% \item \ES{others?}
%
\begin{figure}[t]
\centerline{
\begin{minipage}{1.2\textwidth}
\includegraphics[width=0.5\linewidth]{trace_size_step.png}
\includegraphics[width=0.5\linewidth]{trace_size_jump.png}
\end{minipage}
}
\vspace{3ex}
\caption{Complexity of the generated traces. 81\% of the combined traces
  have a jump complexity of at most 10, with an average complexity of 7
  and a median of 5.}
\label{fig:results-complexity}
\end{figure}
%
\paragraph{Results}
\label{sec:results-complexity}
The results of the experiment are summarized in
Figure~\ref{fig:results-complexity}.
%
The average number of single-step reductions per trace is 31 for the
\ucsdbench\ dataset (35 for the \uwbench\ dataset) with a maximum of
2,745 (986 for \uwbench) and a median of 17 (also 17 for \uwbench).
%
The average number of jumps per trace is 7 (also 7 for \uwbench) with a
maximium of 353 (185 for \uwbench) and a median of 4 (also 4 for
\uwbench).
%
In both datasets 80\% or more traces have at most 10 jumps.

@@

\subparagraph{Construct}
%
We used the correctness of the student's explanation of, and fix for,
the type error as a proxy for her understanding, but it is possible
that other metrics would produce different results.

\subparagraph{Internal}
%
We assigned students randomly to two groups. The first group was given
\ocaml's errors for @append@ and \hbox{@digitsOfInt@,} and \toolname's trace
for @sumList@ and @wwhile@. The second group was given the opposite
assignment of errors and traces. This assignment ensured that (1) each
student was given \ocaml and \toolname problems, and (2) each student
was given an ``easy'' and ``hard'' problem for both \ocaml and
\toolname. Students without sufficient knowledge of \ocaml could affect
the results, as could the time-constrained nature of an exam. For these
reasons we excluded any answers left blank from our analysis.

\subparagraph{External}
%
Our experiment is based on students in the process of learning \ocaml,
and thus may not generalize to all developers. The four
programs we used were chosen manually, via a random selection and
filtering of the programs in the \ucsdbench dataset. In some cases we made
minor simplifying edits (\eg alpha-renaming, dead-code removal) to the
programs to make them more understandable in the short timeframe of an
exam; however, we never altered the resulting type-error. A different
selection of programs may lead to different results.

\subparagraph{Conclusion}
%
We collected exams from 60 students, though due to the nature of the
study not every student completed every problem.
%
The number of complete submissions ranges from 13 (for the \toolname
version of @wwhile@) to 28 (for the \ocaml version of @sumList@), out of
a maximum of 30 per program-tool pair.
%
Collecting more responses per test pair was not possible, as it
would require having students answer the same problem twice (once with
\ocaml and once with \toolname).

@@

To summarize, our experiments demonstrate that \nanomaly finds witnesses
to type errors: (1) with high coverage in a timespan amenable to
compile-time analysis, (2) with traces that have a low average
complexity of 7 jumps, and (3) that are more helpful to novice
programmers than traditional type error messages.

>>>>

\paragraph{Evaluation Goals}
%
There are four questions we seek to answer with our evaluation:
%
\begin{enumerate}
\item \emphbf{Witness Coverage}
      How many ill-typed programs can we find witnesses for?
\item \emphbf{Witness Complexity}
      How \emph{complex} are the traces produced by the witnesses?
\item \emphbf{Witness Utility}
      How \emph{helpful} %(qualitatively and quantitatively)
      are the witnesses and traces in debugging type errors?
\item \emphbf{Witness-based Blame}
      Can witnesses be used to \emph{locate} the source
      of a type error?
\end{enumerate}

@@

We answer the last question on a subset of the \ucsdbench dataset.
%
For each ill-typed program we identify the student's \emph{fix}
by searching for the first type-correct program that follows it in
the interaction trace.
%
We then use an expression-level \emph{diff}~\cite{Lempsink2009-xf} to
determine which sub-expressions changed between the ill-typed program
and the student's fix, and treat those expressions as the source of the
type error.

@@

% \item \ES{others?}
%
\begin{figure}[t]
\centerline{
\begin{minipage}{1.2\textwidth}
\includegraphics[width=0.5\linewidth]{trace_size_step.png}
\includegraphics[width=0.5\linewidth]{trace_size_jump.png}
\end{minipage}
}
\vspace{3ex}
\caption{Complexity of the generated traces. 81\% of the combined traces
  have a jump complexity of at most 10, with an average complexity of 7
  and a median of 5.}
\label{fig:results-complexity}
\end{figure}
%
\paragraph{Results}
\label{sec:results-complexity}
The results of the experiment are summarized in
Figure~\ref{fig:results-complexity}.
%
The average number of single-step reductions per trace is 31 for the
\ucsdbench\ dataset (35 for the \uwbench\ dataset) with a maximum of
2,745 (\resp 986) and a median of 17 (\resp 17).
%
The average number of jumps per trace is 7 (\resp 7) with a
maximium of 353 (\resp 185) and a median of 4 (\resp 4).
%
In both datasets 80\% or more traces have at most 10 jumps.

@@

\subparagraph{Construct.}
%
We used the correctness of the student's explanation of, and fix for,
the type error as a proxy for her understanding, but it is possible
that other metrics would produce different results.

\subparagraph{Internal.}
%
We assigned students randomly to two groups. The first group was given
\ocaml's errors for @append@ and \hbox{@digitsOfInt@,} and \toolname's trace
for @sumList@ and @wwhile@. The second group was given the opposite
assignment of errors and traces. This assignment ensured that (1) each
student was given \ocaml and \toolname problems, and (2) each student
was given an ``easy'' and ``hard'' problem for both \ocaml and
\toolname. Students without sufficient knowledge of \ocaml could affect
the results, as could the time-constrained nature of an exam. For these
reasons we excluded any answers left blank from our analysis.

\subparagraph{External.}
%
Our experiment used students in the process of learning \ocaml,
and thus may not generalize to all developers. The four
programs were chosen manually, via a random selection and
filtering of the programs in the \ucsdbench dataset. In some cases we made
minor simplifying edits (\eg alpha-renaming, dead-code removal) to the
programs to make them more understandable in the short timeframe of an
exam; however, we never altered the resulting type-error. A different
selection of programs may lead to different results.

\subparagraph{Conclusion.}
%
We collected exams from 60 students, though due to the nature of the
study not every student completed every problem.
%
The number of complete submissions ranges from 13 (for the \toolname
version of @wwhile@) to 28 (for the \ocaml version of @sumList@), out of
a maximum of 30 per program-tool pair.
%
Collecting more responses per test pair was not possible, as it
would require having students answer the same problem twice (once with
\ocaml and once with \toolname).

@@

To summarize, our experiments demonstrate that \nanomaly finds witnesses
to type errors: (1) with high coverage in a timespan amenable to
compile-time analysis; (2) with traces that have a low average
complexity of 7 jumps; (3) that are more helpful to novice
programmers than traditional type error messages; and (4) that
can be used to automatically locate the source of a type error.

### 2017-04-27 paper/jfp-submission/interactive-short.tex
prune orphans

In this section we present an interactive visualization of program
executions.
%
First, we extend our semantics (\S~\ref{sec:inter-semant}) to record
each reduction step in a \emph{trace}, producing a \emph{reduction
  graph} alongside the witness.
%
Then we describe a set of common \emph{interactive debugging} steps that
can be expressed as simple traversals over the reduction graph
(\S~\ref{sec:traversing-graph}), yielding an interactive debugger that
allows the user to effectively visualize \emph{how} the program goes
(wrong).

@@

\paragraph{Reduction Graphs}
%
A \emph{steps-to} edge is a pair of expressions \singlestep{e_1}{e_2}, which
intuitively indicates that $e_1$ reduces, in a single step, to $e_2$.
%
%A \emph{sub-term} edge is a pair of expressions \subterm{e_1}{e_2}, which
%intuitively indicates that $e_1$ contains $e_2$ as a sub-expression.
%
A \emph{reduction graph} is a set of steps-to edges:
$$\tr ::= \bullet \spmid \singlestep{e}{e}; \tr$$  %% \spmid \subterm{e}{e}; \tr$$

\paragraph{Tracing Semantics}
%
We extend the transition relation (\S~\ref{sec:semantics}) to
collect the set of edges corresponding to the reduction graph.
%
Concretely, we extend the operational semantics to
a relation of the form $\stepg{e}{\vsu}{\tsu}{\tr}{e'}{\vsu'}{\tsu'}{\tr'}$
where $\tr'$ collects the edges of the transition.

@@

\paragraph{Commands}
Our debugger supports the following \emph{commands}, each of which
is parameterized by a single expression (vertex) selected from the
(current) visualization state:
%
\begin{itemize}
%
\item \stepforwardsym, \stepbackwardsym:
      show the result of a single step forward or backward respectively,
%
\item \jumpforwardsym, \jumpbackwardsym:
      show the result of taking multiple steps (a \emph{``big''} step)
      up to the first function call, or return, forward or backward
      respectively,
%
\item \stepintosym:
      show the result of stepping into a function call in a sub-term,
      isolating it in a new thread, and
%
\item \stepoversym:
      show the result of skipping over a function call in a sub-term.
\end{itemize}

>>>>

In this section we present an interactive visualization of program
executions.
%
First, we extend our semantics (\S~\ref{sec:inter-semant}) to record
each reduction step in a \emph{trace}, producing a \emph{reduction
  graph} alongside the witness.
%
Then we describe a set of common \emph{interactive debugging} steps that
can be expressed as simple traversals over the reduction graph
(\S~\ref{sec:traversing-graph}), yielding an interactive debugger that
allows the user to visualize \emph{how} the program goes
(wrong).

@@

\paragraph{Reduction Graphs}
%
A \emph{steps-to} edge is a pair of expressions \singlestep{e_1}{e_2}, which
indicates that $e_1$ reduces, in a single step, to $e_2$.
%
%A \emph{sub-term} edge is a pair of expressions \subterm{e_1}{e_2}, which
%intuitively indicates that $e_1$ contains $e_2$ as a sub-expression.
%
A \emph{reduction graph} is a set of steps-to edges:
$$\tr ::= \bullet \spmid \singlestep{e}{e}; \tr$$  %% \spmid \subterm{e}{e}; \tr$$

\paragraph{Tracing Semantics}
%
We extend the transition relation (\S~\ref{sec:semantics}) to
collect the set of edges corresponding to the reduction graph.
%
Concretely, we extend the operational semantics to
a relation of the form $\stepg{e}{\vsu}{\tsu}{\tr}{e'}{\vsu'}{\tsu'}{\tr'}$
where $\tr'$ collects the transitions.

@@

\paragraph{Commands}
Our debugger supports the following \emph{commands}, each of which
is parameterized by a single expression (vertex) selected from the
(current) visualization state:
%
\begin{itemize}
%
\item \stepforwardsym, \stepbackwardsym:
      show the result of a single step forward or backward;
%
\item \jumpforwardsym, \jumpbackwardsym:
      show the result of taking multiple steps (a \emph{``big''} step)
      up to the first function call, or return, forward or backward
      respectively;
%
\item \stepintosym:
      show the result of stepping into a function call in a sub-term,
      isolating it in a new reduction thread; and
%
\item \stepoversym:
      show the result of skipping over a function call in a sub-term.
\end{itemize}

### 2017-04-27 paper/jfp-submission/safe.tex
prune orphans

\paragraph{Overloaded Functions.}
%
We found that for 5\% programs \toolname got stuck when it tried to
compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While extremely convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
Consider, for example
%
\begin{code}
  let bad x y =
    if x < y then
      1 + true
    else
      0
\end{code}
%
When given @bad@ as input, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@x < y@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |x|
and |y| such that |x < y| is true.

Extending \toolname with support for symbolic execution would alleviate
this issue, as we could then begin symbolically executing the program
until we learn how to instantiate |x| and |y|.
%
Alternatively, we could \emph{speculatively} instantiate both |x| and
|y| with some arbitrary type, and proceed with execution until we
discover a type error.
%
This speculative instantiation is of course unsound, we would have to
take care to avoid reporting frivolous type errors that were caused by
such instantiations.
%
We would need to track which holes were instantiated speculatively in
order to distinguish type errors that would have happened regardless, as
in the above @bad@ function, from type errors that were the fault of our
instantiation.

@@

Of course, \ocaml cannot be expected to know that this particular call
is safe, its type system is not sophisticated enough to express the
necessary conditions.
%
\ES{TODO: add some comment about using dependent types, though this particular example would be quite bizarre to encode in DT..}

>>>>

\paragraph{Overloaded Functions.}
%
We found that for 5\% programs \toolname got stuck when it tried to
compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
Consider, for example
%
\begin{code}
  let bad x y =
    if x < y then
      1 + true
    else
      0
\end{code}
%
When given @bad@, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@x < y@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |x|
and |y| such that |x < y| is true.

Extending \toolname with support for symbolic execution would alleviate
this issue, as we could then begin symbolically executing the program
until we learn how to instantiate |x| and |y|.
%
Alternatively, we could \emph{speculatively} instantiate both |x| and
|y| with some arbitrary type, and proceed with execution until we
discover a type error.
%
This speculative instantiation is of course unsound, we would have to
take care to avoid reporting frivolous type errors that were caused by
such instantiations.
%
We would need to track which holes were instantiated speculatively in
order to distinguish type errors that would have happened regardless, as
in @bad@, from type errors that were caused by our instantiation.

@@

Of course, \ocaml cannot be expected to know that this particular call
is safe, its type system is not sophisticated enough to express the
necessary conditions.
%
% \ES{TODO: add some comment about using dependent types, though this particular example would be quite bizarre to encode in DT..}

### 2017-04-27 paper/jfp-submission/semantics.tex
prune orphans

The evaluation relation is parameterized by a pair of functions:
called \emph{narrow} (\forcesym) and \emph{generate} (\gensym),
that ``dynamically'' perform type-checking and hole-filling
respectively.

@@

\paragraph{Preservation}
%
We prove two preservation lemmas. First, we show that each evaluation
step refines the partial input type of $f$, thus preserving type
compatibility.
%
\begin{lem}
\label{lem:vsu-ext}
If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and $\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
  (\ie\ $\trace'$ is a single-step extension of $\trace$)
% and $\trace'$ is a single-step extension of $\trace$
and $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$
% and $\resolve{\thole}{\tsu} \neq \resolve{\thole}{\tsu'}$
then $\tsu' = \tsu[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  $\thole$ does not change, so if the partial input types differ then
  $\tsu \neq \tsu'$.
  %
  Note that only \forcesym\ can change \tsu,
  %, and it can only do so via \unifysym
  via \unifysym, which can only extend \tsu.
  % and only the first case,
  % when we extend \vsu\ with $[\ehole \mapsto v]$,
  % does.
  %
  % \ES{fill in cases in the appendix}
\end{proof}
%
Second, we show that at each step of evaluation, the partial input type of $f$
is a refinement of the instantiation of $\vhole{\thole}$.
%
\begin{lem}
\label{lem:resolve-compat}
For all traces
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
$\vsub{\ptype{\trace}{f}}{\resolve{\vhole{\thole}}{\vsu}}$.
\end{lem}
\begin{proof}
  By induction on $\trace$.
  %
  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
  and $\thole$ is trivially a refinement of $\vhole{\thole}$.
  %
  In the inductive case, consider the single-step extension of $\trace$,
  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
  %
  We show by case analysis on the evaluation rules that if
  $\vsub{\ptype{\trace}{f}}{\resolve{\vhole{\thole}}{\vsu}}$, then
  $\vsub{\ptype{\trace'}{f}}{\resolve{\vhole{\thole}}{\vsu'}}$.
  %
\end{proof}

@@

\paragraph{Inputs and Outputs}
%
The problem of generating inputs is undecidable in general.
%
Our witness generation procedure takes two inputs:
%
(1) a search bound $k$ which is used to define the \emph{number} of
traces to explore\footnote{We assume, without loss of generality, that all
traces are finite.} and
%
(2) the target expression $e$ that contains the type error
(which may be a curried function of multiple arguments).
%
The witness generation procedure returns as output a list of (general)
witness expressions, each of which is of the form $e\ v_1 \ldots v_n$.
%
The \emph{empty} list is returned when no witness can be found after
exploring $k$ traces.

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If $\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}
\begin{proof}
  For any function $f$ of multiple arguments, we can define $f'$ to be
  the uncurried version of $f$ that takes all of its arguments as a single
  nested pair, and then apply Theorem~\ref{thm:soundness} to $f'$.
\end{proof}

>>>>

The evaluation relation is parameterized by a pair of functions,
\emph{narrow} (\forcesym) and \emph{generate} (\gensym),
that ``dynamically'' perform type-checking and hole-filling
respectively.

@@

\paragraph{Preservation}
%
We prove two preservation lemmas. First, we show that each evaluation
step refines the partial input type of $f$, thus preserving type
compatibility.
%
\begin{lem}
\label{lem:vsu-ext}
If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and $\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
  (\ie\ $\trace'$ is a single-step extension of $\trace$)
% and $\trace'$ is a single-step extension of $\trace$
and $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$
% and $\resolve{\thole}{\tsu} \neq \resolve{\thole}{\tsu'}$
then $\tsu' = \tsu[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  $\thole$ does not change, so if the partial input types differ then
  $\tsu \neq \tsu'$.
  %
  Only \forcesym\ can change \tsu,
  %, and it can only do so via \unifysym
  via \unifysym, which can only extend \tsu.
  % and only the first case,
  % when we extend \vsu\ with $[\ehole \mapsto v]$,
  % does.
  %
  % \ES{fill in cases in the appendix}
\end{proof}
%
Second, we show that at each step of evaluation, the partial input type of $f$
is a refinement of the instantiation of $\vhole{\thole}$.
%
\begin{lem}
\label{lem:resolve-compat}
For all traces
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
$\vsub{\ptype{\trace}{f}}{\resolve{\vhole{\thole}}{\vsu}}$.
\end{lem}
\begin{proof}
  By induction on $\trace$.
  %
  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
  and $\thole$ trivially refines $\vhole{\thole}$.
  %
  In the inductive case, consider the single-step extension of $\trace$,
  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
  %
  We show by case analysis on the evaluation rules that if
  $\vsub{\ptype{\trace}{f}}{\resolve{\vhole{\thole}}{\vsu}}$, then
  $\vsub{\ptype{\trace'}{f}}{\resolve{\vhole{\thole}}{\vsu'}}$.
  %
\end{proof}

@@

\paragraph{Inputs and Outputs}
%
The problem of generating inputs is undecidable in general.
%
Our witness generation procedure takes two inputs:
%
(1) a search bound $k$ which is used to define the \emph{number} of
traces to explore\footnote{We assume, without loss of generality, that all
traces are finite.} and
%
(2) the target expression $e$ that contains the type error
(which may be a curried function of multiple arguments).
%
The witness generation procedure returns a list of (general)
witness expressions, each of which is of the form $e\ v_1 \ldots v_n$.
%
The \emph{empty} list is returned when no witness can be found after
exploring $k$ traces.

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If $\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}
\begin{proof}
  For any function $f$ of multiple arguments, we can define $f'$ as the
  uncurried version of $f$ that takes all of its arguments as a single
  nested pair, and then apply Theorem~\ref{thm:soundness} to $f'$.
\end{proof}
